# (PART) Workshop Series {-}

# Introduction to R: Part One

```{r 010-setup, include=FALSE}
library(dagitty)
library(magrittr)
```

To learn everything there is to know about R, [read the docs](https://cran.r-project.org/doc/manuals/R-lang.html). To learn just enough to be dangerous, follow along :)

## Introductions and Overview

## Opening the R command line

The R command line interface (CLI) is a *terminal* or console for entering R commands. A *command* is a way of giving the computer some information and having it perform a task.

### Open in Windows

-   start menu (see video below)
-   desktop shortcut
-   task bar shortcut

<video width="100%" autoplay muted loop controls>
  <source src="media/vid/01-opening-r-gui.webm" type="video/webm">
</video>

### Open in Linux

-   application menu
-   dock shortcut
-   terminal (see video below)

<video width="100%" autoplay muted loop controls>
  <source src="media/vid/01-opening-r-cli.webm" type="video/webm">
</video>

## Using the console as a calculator

### Basic arithmetic operators

| Operator | Description      | Example        |
|:---------|:-----------------|:---------------|
| `+`      | addition         | `3 + 4 = 7`    |
| `-`      | subtraction      | `10 - 2 = 8`   |
| `*`      | multiplication   | `2.5 * 4 = 10` |
| `/`      | division         | `9 / 2 = 4.5`  |
| `^`      | exponentiation   | `11^2 = 121`   |
| `%%`     | modulus          | `5 %% 3 = 2`   |
| `%/%`    | integer division | `9 %/% 2 = 4`  |

```{r 010-Navy-Flannel}
# Order of operations still applies
5 - 4 + 5*2^4
# Use parentheses to ensure desired order of operations
(-2 + (5 / 3.8)) * 1.618^2
```

### Three basic types of numbers

I say there are three types of numbers in R because R has different ways of working with different numbers

| Number type | Description                                              | Example        |
|:------------|:---------------------------------------------------------|:---------------|
| Integer     | Whole numbers, positive or negative                      | `4L, -3L, 0L`  |
| Decimal     | Any number, positive or negative, with a fractional part | `3.1, pi, 4.0` |
| Complex     | A real decimal number with an imaginary decimal part     | `-2.1+5i`      |

Notice that the integers have an `L` at the end. This is to explicitly tell R to represent a number as a [[long](https://en.wikipedia.org/wiki/Integer_(computer_science)#Long_integer)] integer, rather than a double (decimal) number.

```{r 010-Global-Toupee}
is.integer(4)  # FALSE
is.double(4)   # TRUE
is.integer(4L) # TRUE
```

### Logical operators

Logical operators test for a condition. They ask a question, and the answer is either `TRUE` or `FALSE`.

| Operator | Description            | Example                                         |
|:---------|:-----------------------|:------------------------------------------------|
| `<`      | less than              | `5 < 4 -> FALSE`                                |
| `<=`     | less than or equal     | `6 <= 6 -> TRUE`                                |
| `>`      | greater than           | `Inf > 1e6 -> TRUE`                             |
| `>=`     | greater than or equal  | `3.5 >= 9 -> FALSE`                             |
| `==`     | exactly equal          | `sin(0) == 0 -> TRUE`                           |
| `!=`     | not equal              | `(3+1)^2 != 3^2 + 1^1 -> TRUE`                  |
| `!`      | not                    | `!TRUE -> FALSE`                                |
| `&&`     | and (short-circuiting) | `(5 < 7) && (7 < 10) -> TRUE`                   |
| `||`     | or (short-circuiting)  | `(2 > 7) || (2 < 4) -> TRUE`                    |
| `&`      | and (vectorized)       | `TRUE & 2 > c(1,2,3) -> c(FALSE, FALSE, TRUE)`  |
| `|`      | or (vectorized)        | `FALSE | 2 <= c(1,2,3) -> c(FALSE, TRUE, TRUE)` |

### Other math functions

| Function  | Description | Example |
|:----------|:------------|:--------|
| `abs`   | absolute value of a number | `abs(-2.5), abs(3L), abs(3+4i)` |
| `sqrt`  | square root of a number    | `sqrt(0.81), sqrt(-1+0i)`       |
| `exp`   | $e^x$                      | `exp(1) ->` `r exp(1)`          |
| `log`   | logarithm of a number (base is the natural number $e$ by default)       | |
| `log10` | base 10 logarithm of a number (see also `log2`, `log1p`, `logb`)        | |
| `sin`, `cos`, `tan` | basic trigonometric functions (radians by default)          | |
| `asin`, `acos`, `atan` | inverse trigonometric functions                          | |
| `sinh`, `asinh`, etc. | hyperbolic and inverse hyperbolic trigonometric functions | |
| `ceiling` | rounds a number towards infinity        | `ceiling(-2.8) -> -2` |
| `floor` | rounds a number towards negative infinity | `floor(3.9) -> 3`     |
| `round` | rounds a number to a specified number of digits (round to even by default) | |
| `trunc` | strips the decimal part from a number | `trunc(2.1) -> 2`, `trunc(2.9) -> 2` |

Also note, all *infix* operations can be used with *prefix* notation:

```{r 010-Sad-Door}
3 + 4 # infix
`+`(3, 4) # prefix
```


### Built-in constants and other values

| Keyword | Description                                           | Example          |
|:--------|:------------------------------------------------------|:-----------------|
| `Inf`   | The floating point representation of infinity         | `1 / 0 -> Inf`   |
| `-Inf`  | Negative infinity                                     | `1 / -0 -> -Inf` |
| `NaN`   | **N**ot **a** **N**umber                              | `0 / 0 -> NaN`   |
| `NA`    | A placeholder for missing values                      |                  |
| `NULL`  | An R object representing the empty set or nothing     |                  |
| `pi`    | The ratio of a circle's circumference to its diameter | `3.14159...`     |

The *Institute of Electrical and Electronics Engineers* (IEEE, "eye-triple-ee") sets many of the standards for computer protocol, including how *floating point* numbers behave. A floating point number is the computer representation of a real number, but due to finite precision, there are some inconsistencies:

```{r 010-Olive-Toupee}
0.3 - 0.1 # should be 0.2
(0.3 - 0.1) == 0.2 # should be true
# R is lying to us!
print(0.3-0.1, digits = 20)
# use `all.equal` to test floating point equality
all.equal(0.3 - 0.1, 0.2)
```

- R uses 64 bit (double) precision for floating point numbers and 32 bit *signed* integers
- https://en.wikipedia.org/wiki/IEEE_754
- https://en.wikipedia.org/wiki/Division_by_zero#Computer_arithmetic

### Practice

The quadratic equation is used to find the roots of quadratic (2nd degree) polynomials. For a quadratic polynomial given by 

$$ax^2 + bx + c$$

the roots are given by

$$x_{1,2} = \frac{-b \pm \sqrt{b^2 -4ac}}{2a}$$

Find the roots of the following equations using the quadratic equation:

a) $y = 2x^2 + 2x - 24$
b) $y = 4x^2 + 9$
c) $y = x^2 + 4x + 4$

## Accessing the built-in docs

Most if not all functions in R have useful help documentation that can be accessed directly within the R console.

``` r
> help(sum)

> ?sum

> library(help = "base")
```

## Working with variables

- Let you store a number or other object in the *environment* list
- Similar to "memory recall" of a calculator, but better

To create a new variable, we can use *left* or *right* assignment:

```{r 010-Balcony Hollow}
# left assignment

# right assignment

```

- You can use an "equal sign" for assignment, but it is discouraged

To list all variables in the session (environment), do:

```{r 010-Sleepy Limousine}
# list all variables in the environment

```

We can see what is stored in those variables by typing them into the console:

```{r 010-Tasty Rhinestone}
# show what is stored in the variables

```

**Good variable naming conventions (in R)**

- See the [tidyverse style guide](https://style.tidyverse.org/syntax.html#object-names) for a full rundown of object naming
- let your names be meaningful
- let your names be consistent
- avoid using existing function names (e.g. `mean`, `c`) and objects (e.g. `T`, `F`, `pi`)
  - List of reserved words in R: https://cran.r-project.org/doc/manuals/R-lang.html#Reserved-words
- avoid using dots
- use the existing conventions of whatever project you are contributing to

## Basic data types in R

```{r 010-Big Hot Cat}
# logical (boolean)

# integer

# double

# complex

# character

```


```{r 010-Needless Storm}
# show the element type

```


## Working with different data types

*Coercion and conversion*

- R will always coerce (convert) an object to the least restrictive type

```{r 010-Flannel Yellow}
typeof(TRUE + 3L)
typeof(TRUE + 3.0)
typeof(TRUE + 3 + 0i)
```

Guess the output type

```{r 010-Elastic Python}
# a "regular" number
14
# integer division
14 %/% 3
# remainder
14 %% 3
# ceiling/floor/round
ceiling(pi)
round(pi)
```
How can we ensure the output type of an operation? Most of the time it's okay to let R worry about the *return* types. In other situations, there's the `as` command:

```{r 010-Late Hammer}
as.integer(4.0)
as.logical(4L) # any non-zero value is true
as.complex(4)
as.double("3.14159")
```

Careful, as not every value can be coerced to every other type, or no robust checks will be performed to ensure validity of a conversion:

```{r 010-Avenue Serious}
# converts to an integer via truncation
as.integer(4.5)
```

*Testing for types*

We can use the `is` command to ask R what the type of an object is:

```{r 010-Rapid Weather}
is.integer(4L)
is.double(3+2i)
is.numeric(4)
is.numeric(4L)
is.numeric(4+2i)
```

Sometimes a thing that looks like an integer is really a double:

```{r 010-Steep Hook}
# actually a double
x <- 4
is.integer(x)
```

How to check that a number is an integer:

```{r 010-Endless Crayon}
# suggestions:

```

## Special data type: Factors

A factor *looks* like a string, but *acts* like an integer (most of the time).

```{r 010-Hungry Roadrunner}
# most often used for categorical observations
pets <- c("cat", "dog", "lizard", "raccoon")
(pet_obs <- sample(pets, 10, TRUE))
# Turn the character vector into a set of factors

```


## Overview of objects and data structures

For a full overview, see the [R documentation on basic types](https://cran.r-project.org/doc/manuals/R-lang.html#Basic-types).

- Vectors
- Matrices and Arrays
- Lists
- Compound objects
  - Factors
  - Data frames

## Working with vectors, matrices, and arrays

### Vectors

- a *vector* is a collection of *same-type* objects
- aka an *atomic* vector
- all elements coerced to the least restrictive type
- typically created using the *concatenate* command, `c()`

```{r 010-Pure Kangaroo, results='hold'}
# a few vectors
c(4.5, 2, 3.8)
1:4
c(TRUE, FALSE, FALSE)
seq(from = 0.1, to = 1.0, by = 0.4)
LETTERS
```

**Question:** What is meant by "least restrictive type"?

```{r 010-Dead Laser}
# mix of integer and logical

# mix of integer and double

# mix of double and complex

# mix of double and character

```

We can access the elements of a vector by using brackets: `[]`. Note that R uses "one-based indexing", meaning that arrays start at $1$, not $0$.

```{r 010-Massive Tungsten}
# accessing a single element from `letters`

# accessing multiple elements using a vector of indices

# accessing a subset using conditionals

# accessing the last element

```


### Matrices

- A matrix in R is a vector with a `dim` *attribute*
- It is stored in memory the same way as a vector is

```{r 010-Scarlet Crayon}
# creating a matrix using the `matrix()` function
matrix(1:12, nrow = 3, ncol = 4)
# filling a matrix with a single value
matrix("m", nrow = 5, ncol = 2)
```

Matrices (and vectors and arrays) are stored in *column-major order*

```{r 010-Gutsy Logbook}
# default: fills in down each column
matrix(1:12, nrow = 3)
# explicitly fill in accross rows first
matrix(1:12, nrow = 3, byrow = TRUE)
```


Elements of a matrix can either be accessed with row-column pairs, or with *linear indexing*

```{r 010-Outstanding Flea}
# create a 3x5 matrix
(M <- matrix(1:15, nrow = 3))
# get the element at the second row and fourth column

# get the 8th element in memory

```

You can create subsets of a matrix using a pair of vectors for indices.

```{r 010-Morbid Obscure}
# get rows 1 and 3, and columns 2 and 4

```

A matrix can be turned back into a vector by removing its dimension attribute. In the same way, a matrix can be reshaped, or a vector can be turned into a matrix.

```{r 010-Serpent Temporary}
# show the attributes of the matrix, M

# get the `dim` attribute

# remove the `dim` attribute by setting it to `NULL`

# turn M into a 5x3 matrix

```

### Arrays

- An array is an extension of a matrix, having 2 or more dimensions
- Arrays are created using the `array()` function or by setting the `dim` attribute of a vector

```{r 010-Magenta Blue Tuba}
# one array
(a <- array(1:(5*4*3), dim = c(5, 4, 3)))
# show the attributes
attributes(a)
```

## Working with lists

- A list is a "generic vector"
- Each element of a list can be *any* type of R object

```{r 010-Outstanding Backpack}
(L <- list(a = 1, "d", formula(y ~ x), sublist = list(a = 1, b = 1:4, c = letters[3:7])))
```

As seen above, lists can hold other lists, and elements can be named or referenced by index. This can become complicated to mentally parse as printed above, so the `str` command can come in handy.

```{r 010-Tidy Boiling Creek}
str(L)
```

Elements of a list can be accessed in three different ways.

```{r 010-Rebel Pluto}
# Using the `$` selector

# Using single bracket notation

# Using double bracket notation

```

## Working with data frames

What are some of the key properties of a data frame?

- 
- 
- 

Accessing elements of a data frame is similar to lists *and* matrices.

```{r 010-Butter Lucky}
# getting a single column using the `$` selector or `[[`

# getting multiple columns using `[` and a index/character vector

# getting a single element using `[`

# getting a single element using `$` and `[`

```

## Reading data from a file

Common text files can be read using the `read-` set of functions

- read comma separated values (CSV) files with `read.csv()`
  - the more general function is `read.delim()`
  - the other specialized function is `read.table()`
- read general text files using `readLines()`

```{r 010-Rough Tombstone}
# read in the sample csv file

```

More proprietary formats can be read in using a library

- `readr` for the tidyverse version of `read-` functions
- `readxl` for Excel spreadsheets
- `R.matlab` for Matlab files

## Fitting a linear model

R has built-in support for fitting [generalized] linear models, with support for numeric and factor variables.

```{r 010-Hungry Boomerang}
# y ~ normal(3x + 2, 0.5)
x <- rnorm(30)
y <- rnorm(length(x), 3*x + 2, 0.5)
plot(y ~ x) # same as plot(x, y)
# fit a linear model

```

Can also fit a generalized linear model using (you guessed it) the `glm()` function.

```{r 010-Swift Electrical}
# logit(theta) = 3x + 2
k <- rbinom(length(x), size = 1, prob = plogis(3*x+2))
plot(k ~ x)
# fit a generalized linear model with logit link

```

## Summarizing data frames and models

R has a method for summarizing vectors, data frames, and fitted models, and it is the `summary()` function.

```{r 010-Skilled Navy Lama}
# summarize a numeric vector

# summarize a data frame

# summarize a linear model

```

Notice that it outputs differently for different input objects. This is because `summary()` is a *generic* function, meaning that it has multiple *methods* defined for different R objects, and packages can define their own behavior for their own types.

## Producing simple plots

There are a few simple plotting functions in R

- `plot()`
  - also "generic"
  - can produce scatter and line plots
- `lines()` adds lines to plots
- `points()` adds points to plots
- `curve()` plots a function over a domain
  - can be used alone or "layered" on top of an existing plot

Other types of graphs include

- `hist()` for histograms
- `barplot()`
- `boxplot()`
- `stem()` for stem-and-leaf plot

## R, compared

*What is R best for*

