```{r 040-setup, include=FALSE, eval=TRUE, cache=FALSE}
options(knitr.kable.NA = '---')

library(kableExtra)
library(tidyverse)
library(lubridate)
```

# Tidy R

## Introductions and Overview

A majority of this workshop is based on the [Tidy Data](https://vita.had.co.nz/papers/tidy-data.pdf) paper by Hadley Wickham and the [corresponding chapter](https://r4ds.had.co.nz/tidy-data.html) in the [R For Data Science](https://r4ds.had.co.nz/) e-book. I highly recommend reading both the paper and the e-book to get a deeper understanding of using R and tidy data.

## Tidy Data

Tidy data is a set of definitions and guidelines to organize data by in order to make it easier to facilitate data analysis. Data can be organized in many different ways, and it may seem like there is either no "best" way to arrange data. Take the following imaginary data. Table  \@ref(tab:040-imaginary-data-1) has two columns and three rows in a commonly seen format:

```{r 040-imaginary-data-1, echo=FALSE, eval=TRUE}
df_patient <- tribble(
  ~`Treatment A`, ~`Treatment B`,
  NA, 2,
  16, 11,
  3, 1
) %>%
  as.data.frame() %>%
  `rownames<-`(c("John Smith", "Jane Doe", "Mary Johnson"))

df_patient %>%
  kable(caption = "Each row has patient data")
```

The same data can also be presented as in table \@ref(tab:040-imaginary-data-2), this time a table with three columns and two rows (a transposition of the first table). Neither is more correct than the other, but necessity may dictate that one is more easy to work with than the other.

```{r 040-imaginary-data-2, eval=TRUE, echo=FALSE}
df_treatment <- df_patient %>%
  as_tibble(rownames = "Name") %>%
  pivot_longer(`Treatment A`:`Treatment B`, names_to = "Treatment", values_to = "Outcome") %>%
  pivot_wider(names_from = Name, values_from = Outcome) %>%
  as.data.frame()

rownames(df_treatment) <- df_treatment$Treatment
df_treatment$Treatment <- NULL
df_treatment %>%
  kable(caption = "Each row is the treatment data")
```

Tidy data defines a few terms to "describe the underlying semantics ... of the values displayed in [a] table."

- **Values**
  - A single element of a data set
  - Every value belongs to a variable and an observation
- **Variables**
  - A collection of all values that measure the same thing
- **Observation**
  - Contains all values measured on the same unit

Consider now the same imaginary data as above, but now in a tidy format:

```{r 040-imaginary-data-3, eval=TRUE, echo=FALSE}
df_tidy <- df_patient %>%
  as_tibble(rownames = "Name") %>%
  pivot_longer(`Treatment A`:`Treatment B`, names_to = "Treatment", values_to = "Outcome") %>%
  mutate(Treatment = stringr::str_remove(Treatment, "Treatment "))

df_tidy %>%
  kable(caption = "Each row is an observation, and each column is a variable.")
```

In **tidy data**:

1) Each variable forms a column
2) Each observation forms a row
3) Each type of observational unit forms a table

Tidy data is now well-defined, and messy data is data that is in any other format. Messy data is not useless, though, and certain analyses and operations require data to be in a format other than tidy. But Tidy data can serve as a launching point (or target) for cleaning, plotting, and analyzing a new data set.

## Tidying Messy Data

Five most common problems of messy data sets:

1) Column headers are values, not variable names
2) Multiple variables are stored in one column
3) Variable names are stored in both rows and columns
4) Multiple types of observational units are stored in the same variable
5) A single observational unit is stored in multiple tables

A small set of tools can be used to clean most of the problems of messy data:

1) Melting (pivot longer)
2) String manipulation (splitting)
3) Casting (pivot wider)

### Data in the column headers

**Messy**

```{r 040-Remote-Pineapple, eval=TRUE, echo=FALSE, warning=FALSE}
if (!file.exists("data/pew.sav")) {
  download.file(
    url = "https://github.com/hadley/tidy-data/raw/master/data/pew.sav",
    destfile = "data/pew.sav"
  )
}
pew <- foreign::read.spss("data/pew.sav")
pew <- as.data.frame(pew)

religion <- pew[c("q16", "reltrad", "income")]
religion$reltrad <- as.character(religion$reltrad)
religion$reltrad <- str_replace(religion$reltrad, " Churches", "")
religion$reltrad <- str_replace(religion$reltrad, " Protestant", " Prot")
religion$reltrad[religion$q16 == " Atheist (do not believe in God) "] <- "Atheist"
religion$reltrad[religion$q16 == " Agnostic (not sure if there is a God) "] <- "Agnostic"
religion$reltrad <- str_trim(religion$reltrad)
religion$reltrad <- str_replace_all(religion$reltrad, " \\(.*?\\)", "")

religion$income <- c("Less than $10,000" = "<$10k", 
  "10 to under $20,000" = "$10-20k", 
  "20 to under $30,000" = "$20-30k", 
  "30 to under $40,000" = "$30-40k", 
  "40 to under $50,000" = "$40-50k", 
  "50 to under $75,000" = "$50-75k",
  "75 to under $100,000" = "$75-100k", 
  "100 to under $150,000" = "$100-150k", 
  "$150,000 or more" = ">150k", 
  "Don't know/Refused (VOL)" = "Don't know/refused")[religion$income]

religion$income <- factor(religion$income, levels = c("<$10k", "$10-20k", "$20-30k", "$30-40k", "$40-50k", "$50-75k", 
  "$75-100k", "$100-150k", ">150k", "Don't know/refused"))

counts <- count(religion, reltrad, income)
names(counts)[1] <- "religion"

pew_raw <- pivot_wider(counts, names_from = income, values_from = n)

pew_raw %>%
  select(all_of(1:7)) %>%
  head(10) %>%
  kable(caption = "Column header with variables")
```

**Tidy**

```{r 040-Lost-Weather, echo=FALSE, eval=TRUE}
counts %>%
  rename(freq = n) %>%
  head(10) %>%
  kable(caption = "Pew data in a tidy format")
```


### Multiple variables in one column

**Messy**

```{r 040-Skilled-Hurricane, echo=FALSE, eval=TRUE}
if (!file.exists("data/tb.csv")) {
  download.file("https://raw.githubusercontent.com/hadley/tidy-data/master/data/tb.csv",
                "data/tb.csv")
}
tb_raw <- read.csv("data/tb.csv", na.strings = "")
tb_raw$new_sp <- NULL
tb_raw <- subset(tb_raw, year == 2000)
names(tb_raw)[1] <- "country"

names(tb_raw) <- str_replace(names(tb_raw), "new_sp_", "")
tb_raw$m04 <- NULL
tb_raw$m514 <- NULL
tb_raw$f04 <- NULL
tb_raw$f514 <- NULL
tb_raw %>%
  as_tibble() %>%
  select(all_of(1:11)) %>%
  head(10) %>%
  kable(caption = "Multiple data in individual columns")
```

**Tidy**

```{r 040-Rebel-Emerald, eval=TRUE, echo=FALSE}
ages <- c(
  "04" = "0-4", 
  "514" = "5-14", 
  "014" = "0-14", 
  "1524" = "15-24", 
  "2534" = "25-34", 
  "3544" = "35-44", 
  "4554" = "45-54", 
  "5564" = "55-64", 
  "65"= "65+", 
  "u" = NA
)

tb_clean <- tb_raw %>%
  pivot_longer(cols = `m014`:`fu`, names_to = "column", values_to = "cases",
               values_drop_na = TRUE) %>%
  arrange(country, column, year) %>%
  separate(column, c("sex", "age"), sep = 1) %>%
  mutate(age = factor(ages[age], levels = ages))

tb_clean %>%
  head(10) %>%
  kable()
```

### Variable names in both rows and columns

**Messy**

```{r 040-Stony-Viper, echo=FALSE, eval=TRUE}
if (!file.exists("data/weather.txt")) {
  download.file(
    "https://raw.githubusercontent.com/hadley/tidy-data/master/data/weather.txt",
    "data/weather.txt"
  )
}

read_fwf2 <- function(path, cols) {
  raw_stations <- readLines(path)
  stations <- data.frame(matrix(ncol = 0, nrow = length(raw_stations)))

  for(i in 1:nrow(cols)) {
    field <- cols[i, ]
    stations[[field$name]] <- stringr::str_trim(
      stringr::str_sub(raw_stations, field$start, field$end))
  }
  stations[stations == ""] <- NA
  stations[] <- lapply(stations, type.convert, as.is = TRUE)
  
  stations
}

cols <- data.frame(
  name =  c("id", "year", "month", "element"),
  start = c(1,     12,    16,      18),
  end =   c(11,    15,    17,      21))

names <- stringr::str_c(c("value", "mflag", "qflag", "sflag"), 
                        rep(1:31, each = 4), 
                        sep = "_")
starts <- cumsum(c(22, rep(c(5, 1, 1, 1), 31)))
starts <- starts[-length(starts)]
ends <- c(starts[-1], starts[length(starts)] + 1) - 1

values <- data.frame(name = names, start = starts, end = ends)
cols <- rbind(cols, values)

# Load data and subset to small example
weather_raw <- read_fwf2("data/weather.txt",  cols)
weather_raw <- subset(weather_raw, year == 2010 & element %in% c("TMIN", "TMAX")) 
weather_raw <- weather_raw[, c(1:4, which(stringr::str_detect(names(weather_raw), "value")))]
weather_raw$id <- str_c(str_sub(weather_raw$id, 1, 2), str_sub(weather_raw$id, -5, -1))

names(weather_raw)[-(1:4)] <- str_c("d", 1:31)
weather_raw[weather_raw == -9999] <- NA
weather_raw[-(1:4)] <- weather_raw[-(1:4)] / 10 
rownames(weather_raw) <- NULL
weather_raw$element <- tolower(weather_raw$element)

weather_raw %>%
  head(10) %>%
  kable()
```

**Tidy**

```{r 040-Old-Postal, echo=FALSE, eval=TRUE}
weather_clean <- weather_raw %>%
  pivot_longer(cols = d1:d31, names_to = "day", values_to = "value", values_drop_na = TRUE) %>%
  mutate(day = stringr::str_remove(day, "d"),
         day = as.integer(day),
         date = as.Date(ISOdate(year, month, day))) %>%
  select(-c(year, month, day)) %>%
  pivot_wider(id_cols = c(id, date), names_from = element, values_from = value)

weather_clean %>%
  head(10) %>%
  kable()
```

### Multiple types of observational units are stored in the same variable

**Messy**

```{r 040-Venus-Digital, echo=FALSE, eval=TRUE}
if (!file.exists("data/billboard.csv")) {
  download.file(
    "https://raw.githubusercontent.com/hadley/tidy-data/master/data/billboard.csv",
    "data/billboard.csv"
  )
}

billboard_raw <- read_csv("data/billboard.csv",
                col_types = cols(time = col_character())) %>%
  select(year, artist.inverted, track, time, date.entered, x1st.week:x76th.week) %>%
  rename(artist = artist.inverted) %>%
  mutate(artist = iconv(artist, "MAC", "ASCII//translit"),
         time = lubridate::ms(time),
         track = stringr::str_replace(track, " \\(.*?\\)", "")) %>%
  arrange(year, artist, track)

names(billboard_raw)[-c(1:5)] <- str_c("wk", 1:76)

billboard_raw %>%
  select(all_of(1:8)) %>%
  head(10) %>%
  kable()
```

**Half Tidy**

```{r 040-Flying-Vegetable, echo=FALSE, eval=TRUE}
billboard_raw_tidy <- billboard_raw %>%
  pivot_longer(cols = wk1:wk76, names_to = "week", values_to = "rank", values_drop_na = TRUE) %>%
  mutate(week = str_remove(week, "wk"),
         week = as.integer(week))

billboard_raw_tidy %>%
  head(10) %>%
  kable()
```

**Tidy**

```{r 040-Boiling-Omega, echo=FALSE, eval=TRUE}
song <- billboard_raw_tidy %>%
  select(artist, track, time) %>%
  distinct() %>%
  mutate(id = row_number()) %>%
  relocate(id, .before = 1)

rank <- left_join(billboard_raw_tidy, song) %>%
  relocate(id, .before = 1) %>%
  select(id, date.entered, rank)

t1 <- kable(head(song, 10))
t2 <- kable(head(rank, 17))
knitr::kables(list(t1, t2))
```

### A single observational unit is stored in multiple tables

It is common to find data values for individual observational units spread out over multiple tables or files. An example is the US census data, where each data set is split up by the year the census was conducted (e.g. `census_2020`, `census_2010`, etc.). 

## Formatting to Tidy Data

### Melting/Pivoting Longer/Stacking

```{r 040-Massive-Monkey, eval=TRUE}
table4a # cases
table4b # population
```

### Joins

```{r 040-Pink-Clown, eval=TRUE}
tidy4a <- table4a
tidy4b <- table4b
# join
```

See the [dplyr cheatsheet](https://www.rstudio.com/resources/cheatsheets/) for a more visual depiction for the various types of joins.

- `bind_rows()` (Base equivalent is `rbind()`)
- `bind_cols()` (Base equivalent is `cbind()`)
- `left_join()`, `right_join()`
- `full_join()`, `inner_join()`

### Casting/Pivoting Wider/Unstacking

```{r 040-Crystal-Stony, eval=TRUE}
table2
```

### String Manipulation

`separate()` (the opposite is `unite()`)

```{r 040-Brutal-Flea, eval=TRUE}
table3
```

`package:stringr`

The `stringr` package is a collection of consistent functions that make operations on strings easier. It is largely a wrapper for the `stringi` package.

```{r 040-Lonesome-Tuba, eval=TRUE}
ls("package:stringr")
```

## Manipulating Tidy Data

Select, transform, summarize

### Group by and Summarize

```{r 040-Surreal-Clown, eval=TRUE}
library(gapminder)
glimpse(gapminder)
```


## Minimizing Repetition

### `package:purrr`

`purrr`: part of the tidyverse, the basic function `map()` is very similar to the `apply` family of functions

- The goal is to minimize repetition, while maximizing replication
- Keep your code D.R.Y.
  
The basic arguments for `map()` are:

- `.x`: A list or vector
- `.f`: A function or formula

```{r 040-Magenta-Metaphor, eval=TRUE}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
# notice you do not need the () after the function as you would if executing from consol
map(df, mean)
# can pipe it an input to make the process even easier
df %>% map(mean)
```

You may have noticed that the output format is not necessarily in the format that we might expect. The `map()` function always outputs a list.

There are helper functions that are explicit about the return types of a map function: `map_lgl()`, `map_int()`, `map_dbl()` and `map_chr()` return an atomic vector of the indicated type (or die trying).

```{r 040-Beacon-Fox, eval=TRUE}
# Should return a numerical vector rather than a list
map_dbl(df, mean)
```

#### Practice

1) What does `map(-2:2, rnorm, n = 5)` do?
2) What about `map_dbl(-2:2, rnorm, n = 5)`? Why?
3) How do we interpret the error coming from running the above code?

## Personal Data Sets

It's time for you all to share your own personal data sets and let's see if we can get them into a Tidy format.
