```{r 040-setup, include=FALSE, eval=TRUE}
library(kableExtra)
library(tidyverse)
library(lubridate)
library(gapminder)
```

# Tidy R

## Introductions and Overview

A majority of this workshop is based on the [Tidy Data](https://vita.had.co.nz/papers/tidy-data.pdf) paper by Hadley Wickham and the [corresponding chapter](https://r4ds.had.co.nz/tidy-data.html) in the [R For Data Science](https://r4ds.had.co.nz/) e-book. I highly recommend reading both the paper and the e-book to get a deeper understanding of using R and tidy data.

## Tidy Data

Tidy data is a set of definitions and guidelines to organize data by in order to make it easier to facilitate data analysis. Data can be organized in many different ways, and it may seem like there is either no "best" way to arrange data. Take the following imaginary data. Table  \@ref(tab:040-imaginary-data-1) has two columns and three rows in a commonly seen format:

```{r 040-imaginary-data-1, echo=FALSE, eval=TRUE}
df_patient <- tribble(
  ~`Treatment A`, ~`Treatment B`,
  NA, 2,
  16, 11,
  3, 1
) %>%
  as.data.frame() %>%
  `rownames<-`(c("John Smith", "Jane Doe", "Mary Johnson"))

df_patient %>%
  kable(caption = "Each row has patient data")
```

The same data can also be presented as in table \@ref(tab:040-imaginary-data-2), this time a table with three columns and two rows (a transposition of the first table). Neither is more correct than the other, but necessity may dictate that one is more easy to work with than the other.

```{r 040-imaginary-data-2, eval=TRUE, echo=FALSE}
df_treatment <- df_patient %>%
  as_tibble(rownames = "Name") %>%
  pivot_longer(`Treatment A`:`Treatment B`, names_to = "Treatment", values_to = "Outcome") %>%
  pivot_wider(names_from = Name, values_from = Outcome) %>%
  as.data.frame()

rownames(df_treatment) <- df_treatment$Treatment
df_treatment$Treatment <- NULL
df_treatment %>%
  kable(caption = "Each row is the treatment data")
```

Tidy data defines a few terms to "describe the underlying semantics ... of the values displayed in [a] table."

- **Values**
  - A single element of a data set
  - Every value belongs to a variable and an observation
- **Variables**
  - A collection of all values that measure the same thing
- **Observation**
  - Contains all values measured on the same unit

Consider now the same imaginary data as above, but now in a tidy format:

```{r 040-imaginary-data-3, eval=TRUE, echo=FALSE}
df_tidy <- df_patient %>%
  as_tibble(rownames = "Name") %>%
  pivot_longer(`Treatment A`:`Treatment B`, names_to = "Treatment", values_to = "Outcome") %>%
  mutate(Treatment = stringr::str_remove(Treatment, "Treatment "))

df_tidy %>%
  kable(caption = "Each row is an observation, and each column is a variable.")
```

In **tidy data**:

1) Each variable forms a column
2) Each observation forms a row
3) Each type of observational unit forms a table

Tidy data is now well-defined, and messy data is data that is in any other format. Messy data is not useless, though, and certain analyses and operations require data to be in a format other than tidy. But Tidy data can serve as a launching point (or target) for cleaning, plotting, and analyzing a new data set.

## Tidying Messy Data

Five most common problems of messy data sets:

1) Column headers are values, not variable names
2) Multiple variables are stored in one column
3) Variable names are stored in both rows and columns
4) Multiple types of observational units are stored in the same variable
5) A single observational unit is stored in multiple tables

A small set of tools can be used to clean most of the problems of messy data:

1) Melting (pivot longer)
2) String manipulation (splitting)
3) Casting (pivot wider)

### Data in the column headers

**Messy**

```{r 040-Remote-Pineapple, eval=TRUE, echo=FALSE, warning=FALSE}
if (!file.exists("data/pew.sav")) {
  download.file(
    url = "https://github.com/hadley/tidy-data/raw/master/data/pew.sav",
    destfile = "data/pew.sav"
  )
}
pew <- foreign::read.spss("data/pew.sav")
pew <- as.data.frame(pew)

religion <- pew[c("q16", "reltrad", "income")]
religion$reltrad <- as.character(religion$reltrad)
religion$reltrad <- str_replace(religion$reltrad, " Churches", "")
religion$reltrad <- str_replace(religion$reltrad, " Protestant", " Prot")
religion$reltrad[religion$q16 == " Atheist (do not believe in God) "] <- "Atheist"
religion$reltrad[religion$q16 == " Agnostic (not sure if there is a God) "] <- "Agnostic"
religion$reltrad <- str_trim(religion$reltrad)
religion$reltrad <- str_replace_all(religion$reltrad, " \\(.*?\\)", "")

religion$income <- c("Less than $10,000" = "<$10k", 
  "10 to under $20,000" = "$10-20k", 
  "20 to under $30,000" = "$20-30k", 
  "30 to under $40,000" = "$30-40k", 
  "40 to under $50,000" = "$40-50k", 
  "50 to under $75,000" = "$50-75k",
  "75 to under $100,000" = "$75-100k", 
  "100 to under $150,000" = "$100-150k", 
  "$150,000 or more" = ">150k", 
  "Don't know/Refused (VOL)" = "Don't know/refused")[religion$income]

religion$income <- factor(religion$income, levels = c("<$10k", "$10-20k", "$20-30k", "$30-40k", "$40-50k", "$50-75k", 
  "$75-100k", "$100-150k", ">150k", "Don't know/refused"))

counts <- count(religion, reltrad, income)
names(counts)[1] <- "religion"

raw <- pivot_wider(counts, names_from = income, values_from = n)

raw %>%
  select(all_of(1:7)) %>%
  head(10) %>%
  kable(caption = "Column header with variables")
```

**Tidy**

```{r 040-Lost-Weather, echo=FALSE, eval=TRUE}
counts %>%
  rename(freq = n) %>%
  head(10) %>%
  kable(caption = "Pew data in a tidy format")
```


### Multiple variables in one column

**Messy**

```{r 040-Skilled-Hurricane, echo=FALSE, eval=TRUE}
if (!file.exists("data/tb.csv")) {
  download.file("https://raw.githubusercontent.com/hadley/tidy-data/master/data/tb.csv",
                "data/tb.csv")
}
raw <- read.csv("data/tb.csv", na.strings = "")
raw$new_sp <- NULL
raw <- subset(raw, year == 2000)
names(raw)[1] <- "country"

names(raw) <- str_replace(names(raw), "new_sp_", "")
raw$m04 <- NULL
raw$m514 <- NULL
raw$f04 <- NULL
raw$f514 <- NULL
raw %>%
  as_tibble() %>%
  select(all_of(1:11)) %>%
  head(10) %>%
  kable(caption = "Multiple data in individual columns")
```

**Tidy**

```{r 040-Rebel-Emerald, eval=TRUE, echo=FALSE}
ages <- c(
  "04" = "0-4", 
  "514" = "5-14", 
  "014" = "0-14", 
  "1524" = "15-24", 
  "2534" = "25-34", 
  "3544" = "35-44", 
  "4554" = "45-54", 
  "5564" = "55-64", 
  "65"= "65+", 
  "u" = NA
)


clean <- raw %>%
  pivot_longer(cols = `m014`:`fu`, names_to = "column", values_to = "cases",
               values_drop_na = TRUE) %>%
  arrange(country, column, year) %>%
  separate(column, c("sex", "age"), sep = 1) %>%
  mutate(age = factor(ages[age], levels = ages))

clean %>%
  head(10) %>%
  kable()
```

### Variable names in both rows and columns

**Messy**

```{r 040-Stony-Viper, echo=FALSE, eval=TRUE}
if (!file.exists("data/weather.txt")) {
  download.file(
    "https://raw.githubusercontent.com/hadley/tidy-data/master/data/weather.txt",
    "data/weather.txt"
  )
}

read_fwf2 <- function(path, cols) {
  raw_stations <- readLines(path)
  stations <- data.frame(matrix(ncol = 0, nrow = length(raw_stations)))

  for(i in 1:nrow(cols)) {
    field <- cols[i, ]
    stations[[field$name]] <- stringr::str_trim(
      stringr::str_sub(raw_stations, field$start, field$end))
  }
  stations[stations == ""] <- NA
  stations[] <- lapply(stations, type.convert, as.is = TRUE)
  
  stations
}

cols <- data.frame(
  name =  c("id", "year", "month", "element"),
  start = c(1,     12,    16,      18),
  end =   c(11,    15,    17,      21))

names <- stringr::str_c(c("value", "mflag", "qflag", "sflag"), 
                        rep(1:31, each = 4), 
                        sep = "_")
starts <- cumsum(c(22, rep(c(5, 1, 1, 1), 31)))
starts <- starts[-length(starts)]
ends <- c(starts[-1], starts[length(starts)] + 1) - 1

values <- data.frame(name = names, start = starts, end = ends)
cols <- rbind(cols, values)

# Load data and subset to small example
raw <- read_fwf2("data/weather.txt",  cols)
raw <- subset(raw, year == 2010 & element %in% c("TMIN", "TMAX")) 
raw <- raw[, c(1:4, which(stringr::str_detect(names(raw), "value")))]
raw$id <- str_c(str_sub(raw$id, 1, 2), str_sub(raw$id, -5, -1))

names(raw)[-(1:4)] <- str_c("d", 1:31)
raw[raw == -9999] <- NA
raw[-(1:4)] <- raw[-(1:4)] / 10 
rownames(raw) <- NULL
raw$element <- tolower(raw$element)

raw %>%
  head(10) %>%
  kable()
```

**Tidy**

```{r 040-Old-Postal, echo=FALSE, eval=TRUE}
clean <- raw %>%
  pivot_longer(cols = d1:d31, names_to = "day", values_to = "value", values_drop_na = TRUE) %>%
  mutate(day = stringr::str_remove(day, "d"),
         day = as.integer(day),
         date = as.Date(ISOdate(year, month, day))) %>%
  select(-c(year, month, day)) %>%
  pivot_wider(id_cols = c(id, date), names_from = element, values_from = value)

clean %>%
  head(10) %>%
  kable()
```

### Multiple types of observational units are stored in the same variable

**Messy**

```{r 040-Venus-Digital, echo=FALSE, eval=TRUE}
if (!file.exists("data/billboard.csv")) {
  download.file(
    "https://raw.githubusercontent.com/hadley/tidy-data/master/data/billboard.csv",
    "data/billboard.csv"
  )
}

raw <- read_csv("data/billboard.csv",
                col_types = cols(time = col_character())) %>%
  select(year, artist.inverted, track, time, date.entered, x1st.week:x76th.week) %>%
  rename(artist = artist.inverted) %>%
  mutate(artist = iconv(artist, "MAC", "ASCII//translit"),
         time = lubridate::ms(time),
         track = stringr::str_replace(track, " \\(.*?\\)", "")) %>%
  arrange(year, artist, track)

names(raw)[-c(1:5)] <- str_c("wk", 1:76)

raw %>%
  select(all_of(1:8)) %>%
  head(10) %>%
  kable()
```

**Half Tidy**

```{r 040-Flying-Vegetable, echo=FALSE, eval=TRUE}
raw_tidy <- raw %>%
  pivot_longer(cols = wk1:wk76, names_to = "week", values_to = "rank", values_drop_na = TRUE) %>%
  mutate(week = str_remove(week, "wk"),
         week = as.integer(week))

raw_tidy %>%
  head(10) %>%
  kable()
```

**Tidy**

```{r 040-Boiling-Omega, echo=FALSE, eval=TRUE}
song <- raw_tidy %>%
  select(artist, track, time) %>%
  distinct() %>%
  mutate(id = row_number()) %>%
  relocate(id, .before = 1)

rank <- left_join(raw_tidy, song) %>%
  relocate(id, .before = 1) %>%
  select(id, date.entered, rank)

t1 <- kable(head(song, 10))
t2 <- kable(head(rank, 17))
knitr::kables(list(t1, t2))
```

### A single observational unit is stored in multiple tables

Common to find data values for individual observational units spread out over multiple tables or files. An example is the US census data, where each data set is split up by the year the census was conducted (e.g. `census_2020`, `census_2010`, etc.). 

## Formatting Tidy Data

### Melting (Pivot Longer)

### String Manipulation

- `separate()`
- `mutate()`
- `stringr`

### Casting (Pivot Wider)

## Manipulating Tidy Data

Select, transform, summarize

### Group by and Summarize

## Minimizing Repetition

### Purrr

purrr: part of the tidyverse, the basic function `map()` is very similar to the `apply` family of functions.
  - The goal is to minimize repetition, while maximizing replication. 
  - Keep your code D.R.Y.
  
The basic arguments for `map()` are:
`.x` - A list or vector (can be logical,integer,numeric,character)
`.f` - A function or formula

```{r 040-MooCow}
# the rnorm function is handy for making test datasets. What does rnorm do?
?rnorm()

library(tidyverse)
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
mn <- map_dbl(df, mean)
# notice you do not need the () after the function as you would if executing from console

# can pipe it an input to make the process even easier
df %>% map_dbl(mean)
```

Can't we just use our function on the character vector though?

```{r 040-MooCow2}
mean(df)

# No silly!
```

PRACTICE: Compute the mean, median, and sd of every column in the mtcars data.

MORE PRACTICE: Incorporate what we have learned in earlier workshops and create a function to output the mean, median and sd of the mtcars data and save it as a new variable.

EVEN MORE PRACTICE: If you are tracking these changes in a git repo, commit them to your github.



When viewing the ouput of the function that we applied to our dummy data, you may have noticed that while the values are correct, the ouput format is messy.
  - the base `map()` function will always output a list
  - what about if we want a numeric outcome? or a list of `chr` values that we can plot later?
  - `map_lgl()`, `map_int()`, `map_dbl()` and `map_chr()` return an atomic vector of the indicated type (or die trying).

Let's see how this works:

```{r 040-Lollipops}

sr_dbl <- map_dbl(df, sqrt)
# Should return a numerical vector rather than a list
view(sr_dbl)

```

What does map(-2:2, rnorm, n = 5) do? 

What about map_dbl(-2:2, rnorm, n = 5)? Why?

How do we interpret this error code?


Pretty Cool! But we can be even more efficient.

### Nested Data

What is nested data? If you know about lists, you are halfway there.

A nested data frame is just a normal data frame, but with more features. It still has columns and rows, but the content of the cells can be:
  - lists
  - models
  - other data frames
  - plots
When combined with the `map()` functions, we can manipulate a large amount of data very quickly. 

Now....what does this look like? We'll do some examples using the `gapminder` dataset:

```{r 040-Elephant Ear}
# Take a peak at the dataset
head(gapminder)

gapminder_nested <- gapminder %>% 
  group_by(continent) %>% 
  nest()

gapminder_nested
```

The first column is the variable that we grouped by, continent, and the second column is the rest of the data frame corresponding to that group (as if you had filtered the data frame to the specific continent). To see this, the code below shows that the first entry in the data column corresponds to the entire gapminder dataset for Asia.

```{r 040-gapminder}
gapminder_nested$data[[1]]
```

So why nest a data frame? It doesn't seem very useful...until you realize you can use other dplyr functions to manipulate objects that can be stored in a list. 

However, many common functions like `mutate()` are applied directly to an entire column (usually a vector), there are problems trying to use these on a list. 

```{r 040-tutoreos}
gapminder_nested %>% 
  mutate(avg_lifeExp = mean(data$lifeExp))
```

I wanted to extract the life expectancy column from each data frame and create a new one. Why doesn't this work? (Hint: open the nested dataset and look at the column types)

Basically, we need to map the `mean()` function to our nested data. We will first try to do it for just one continent with the `pluck()` function, and then for the whole nested dataset:

```{r 040-Guitar-Hero}
asia <- gapminder_nested %>% pluck('data', 1)
# Average life expectancy of Asia could be calculated by:
mean(asia$lifeExp)



gapminder_nested %>% 
  mutate(avg_lifeExp = map_dbl(data, ~{mean(.$lifeExp)}))
```

Bonus: What other functions could we have used to do this?


## Merging Tidy Data

### Bind

### Join







