---
title: "Basics of R, RStudio, and Setting Up the Environment"
format: html
editor: visual
---

Estimated time: 45 minutes

# What is R

R is a programming language focused on statistical computing and graphics. It is commonly used for:

-   Statistical modeling
-   Data analysis and visualization
-   Exploratory and reproducible research
-   Creating high-quality, publication-ready documents

Regarding the last point, `R` can be combined with other tools to produce websites, presentations, books, and other types of documents:

-   [The R for Data Science book](https://r4ds.had.co.nz/)
-   [The Nevada INBRE site is written in R](https://nevadainbre.github.io/)
-   [My thesis was written in R](https://adknudson.github.io/UNR-Masters-Thesis/)

# Installing R and RStudio

## Installing R

We can acquire `R` from the R Project website: <https://www.r-project.org/>

1.  Choose a "mirror" site to download from

-   `0-Cloud` is fine

2.  Select the download link for your operating system

-   `macOS`, `Windows`, or `Linux`

3.  Follow the install instructors for your OS

### Check your installation

Run R in the terminal interface:

`<steps are different for different operating systems>`

You should see a start-up message saying which version of R you are running.

```         
R version 4.4.2 (2024-10-31) -- "Pile of Leaves"
Copyright (C) 2024 The R Foundation for Statistical Computing
Platform: x86_64-pc-linux-gnu
```

You can run some simple commands like elementary operations:

```{r math-basics}
#| eval: false
8 + 4
1 - 2
6 * 7
5 / 3
9^2
```

or other mathematical functions:

```{r math-funs}
#| eval: false
sqrt(2)
sin(pi / 4)
log2(32)
exp(1)
```

## Installing RStudio

RStudio can be downloaded from the posit website: <https://posit.co/download/rstudio-desktop/>

Find the download four your operating system and install in the normal way for your OS.

# Overview of RStudio

See the user guide for a more in-depth walk-through: <https://docs.posit.co/ide/user/>

## Pane Layout

Four primary pains:

-   Source pane
-   Console pane
-   Environment pane
-   Output pane

The layout can be modified by opening the `Tools > Global Options > pane layout`

### Source

-   The source pane is where you can view and edit text-based documents and source code
-   Multiple documents can be opened in separate tabs of the source pane
-   New source windows can be created in two ways
    -   `Show in New Window` button
    -   Drag a source tab onto the desktop

### Console

-   The console pane is an area where executed code output is displayed

-   You can run commands here to test behavior or prototype code

    -   However, it is recommended to prototype code in a script that can be saved

### Environment

-   Displays objects in the current R session

-   Stores recent commands that have been executed

-   Version control system

-   Build tab for building the current project

### Output

-   File browser

-   Plot output

-   List of installed packages

-   Help tab

-   Viewers for document output

## Menu Bar

-   Typical menus: `File`, `Edit`, `View`, `Tools`, `Help`

-   Programming menus: `Code`, `Plots`, `Session`, `Build`, `Debug`, `Profile`

## Settings

`Tools > Global Options`

## Shortcuts

| Name | Shortcut | Description |
|----|----|----|
| Command Palette | `Ctrl + Shift + P` | Opens the command palette to quickly run commands |
| Build Project | `Ctrl + Shift + B` | Builds the current project if it is supported |
| Run Current Chunk | `Ctrl + Shift + Enter` | Runs the entire code chunk at the cursor location |
| Run Current Line | `Ctrl + Enter` | Runs the line at the cursor location |

: Useful Shortcuts

# Creating and Saving R Scripts

-   Create a new script

-   Add code

-   Save the script

-   Close and re-open the script

# Installing Packages

*\<what is a package\>*

Two ways to install packages

**Via the console**

``` r
install.package("packageName")
```

**Via RStudio**

`Tools > Install Packages`

## Troubleshooting

-   Sometimes a package requires external libraries to function properly

-   Check the console output to see if there are any helpful messages

-   Search online to see if there are detailed steps for installation

Anecdotally, installing packages like the `Tidyverse` in Linux requires installing many external libraries. Windows seems to be better about packaging the external libraries with the package you are installing.
