---
title: "Hands-On Basics"
format: html
number-sections: true
editor: visual
execute:
  eval: false
---

# Arithmetic

Most (if not all) programming languages support basic arithmetic operations.

| Operator | Description                 |
|----------|-----------------------------|
| `+`      | Addition                    |
| `-`      | Subtraction                 |
| `*`      | Multiplication              |
| `/`      | Division                    |
| `^`      | Exponentiation              |
| `%/%`    | Integer Division (Quotient) |
| `%%`     | Modulus (Remainder)         |

: Arithmetic Operators

**Basic Operations**

```{r arithmetic-operators}
1 + 2
4 - 3
5 * 6
7 / 8
10^9
```

*Notice the spacing around the operations...*

**Special Operations**

Integer Division (Quotient)

```{r integer-devision-operator}
21 %/% 5
113 %/% 7
```

Take care using negative numbers

```{r integer-division-negative-numbers}
11 %/% 5
11 %/% -5
-11 %/% -5
-11 %/% 5
```

Modulus (Remainder)

```{r modulus-operator}
12 %% 5
31 %% 2
78 %% 2
```

Take care using negative numbers

```{r modulus-negative-numbers}
10 %% 7
10 %% -7
-10 %% -7
-10 %% 7
```

## Exercises

1.  The density of a type of soil is $1385\textrm{ kg}/\textrm{m}^3$. How many kilograms of soil is needed to fill up a garden box that is $0.5\textrm{m} \times 2.5\textrm{m} \times 1\textrm{m}$?

```{r ex1.1}
```

2.  The quadratic formula is used to find the roots of a 2nd degree polynomial.

The quadratic formula:

$$
x_{1,2} = \frac{-b \pm \sqrt{b^2 - 4ac}}{2a}
$$

Find the roots of the polynomial $3x^2 + 4x - 7$

```{r ex1.2}
```

3.  Continuous interest can be calculated with the formula

    $$
    P(t) = P_0 e^{rt}
    $$

Where $P(t)$ is the amount owed after $t$ years, $e$ is Euler's number, and $r$ is the annual interest rate.

What is the total amount owed after $2$ years on $\$500$ loan with an annual interest rate of $4.5\%$?

```{r ex1.3}
```

# Mathematical Functions

| Function | Description |
|:---|:---|
| `abs` | absolute value of a number |
| `sqrt` | square root of a number |
| `exp` | $e^x$ |
| `log` | logarithm of a number (base is the natural number $e$ by default) |
| `log10` | base 10 logarithm of a number (see also `log2`, `log1p`, `logb`) |
| `sin`, `cos`, `tan` | basic trigonometric functions (radians by default) |
| `asin`, `acos`, `atan` | inverse trigonometric functions |
| `sinh`, `asinh`, etc. | hyperbolic and inverse hyperbolic trigonometric functions |
| `ceiling` | rounds a number towards infinity |
| `floor` | rounds a number towards negative infinity |
| `round` | rounds a number to a specified number of digits (round to even by default) |
| `trunc` | strips the decimal part from a number |

```{r math-funs}
abs(-4)
```

# Special Numbers

Math with real numbers follows the [IEEE 754](https://en.wikipedia.org/wiki/IEEE_754) standard for *floating point arithmetic*.

| Number | Description                                               |
|:-------|:----------------------------------------------------------|
| `±Inf` | Floating point representation of positive infinity        |
| `NaN`  | Floating point representation of **N**ot **a** **N**umber |

```{r special-floats}
1 / 0
Inf + 1

-1 / 0
-Inf * 5

Inf - Inf
NaN * 3
```

# Special Constants

| Constant | Description                               |
|----------|-------------------------------------------|
| `pi`     | The circle constant, $3.14159\ldots$      |
| `NA`     | **N**ot **A**vailable, or a missing value |
| `NULL`   | Represents an empty object                |

```{r special-constants}
?pi
?NA
?NULL
```

# Logical Operators

Logical operators test the "truthiness" of an expression.

| Operator | Description            |
|:---------|:-----------------------|
| `<`      | less than              |
| `<=`     | less than or equal     |
| `>`      | greater than           |
| `>=`     | greater than or equal  |
| `==`     | exactly equal          |
| `!=`     | not equal              |
| `!`      | not                    |
| `&&`     | and (short-circuiting) |
| `||`     | or                     |
| `&`      | and (vectorized)       |
| `|`      | or (vectorized)        |

```{r logical-operators}
5 > 3
10 <= 7
6 == 7
-0 != 0
```

# Getting Help

You can access the built-in documentation in several ways:

-   `help(sum)`

-   `?sum`

-   `library(help = "base")`

-   Using the `Help` tab in the viewer pane (lower right)

# Working with Variables

Variables are a way to save intermediate results of a calculation, or to "bind" an object to a name.

Assignment is done with the `<-` operator. This is called "left assignment".

```{r left-assignment}
x <- 1
y <- 2
```

Variables can also be assigned to names on the right via the "right assignment" operator, `->.`

```{r right-assignment}
7 - 2 -> z
```

Right assignment is not as common, but can be more intuitive when working with long chains of operations using the *pipe operator*. You will see an example piping operations in the second half of this workshop.

When an *object* is bound to a name, the object can be accessed by the name it is bound to.

```{r using-variables}
x <- 4
y <- 3
x + y
```

## Variable Name Rules

There are rules to what is considered a *valid identifier.*

-   Variables in `R` can start with lowercase or uppercase letters, or a period

-   Variables *may not* start with a number or any other type of symbol

-   Variables can consist of letters, numbers, underscores, or periods.

**Valid Identifiers**

```{r valid-identifiers}
a <- 4
B <- 76.4
a1b2 <- 6

.myVariable <- 0
my_variable <- 653
my.variable <- 73
```

**Invalid Identifiers**

```{r invalid-identifiers}
_x <- 2
1book <- 10
```

### Good Variable Naming

-   See the [tidyverse style guide](https://style.tidyverse.org/syntax.html#object-names) for a full rundown of object naming
-   Let your names be meaningful
-   Let your names be consistent
-   Avoid using existing function names (e.g. `mean`, `c`) and objects (e.g. `T`, `F`, `pi`)
    -   List of reserved words in R: <https://cran.r-project.org/doc/manuals/R-lang.html#Reserved-words>
-   Avoid using dots
    -   Names with dots are more common for internal variables of a package
-   Use the existing conventions of whatever project you are contributing to

## Listing Variables in the Environment

You can get a list of variables in the environment in two ways:

-   Calling `ls()` in the console

-   Viewing the list of variables in the `Environment` tab (upper right pane)

```{r list-of-variables}
ls()
```

# Data Types

Every object in `R` has a "type"

## Atomic Types

There are a few *atomic* types in the `R` programming language: `logical`, `integer`, `double`, and `character`. These are types that contain only one type of data, and cannot be further divided into smaller components.

In addition to the 4 common atomic types, there are also two other less common atomic types: `complex` and `raw`.

```{r common-atomic-types}
TRUE   # logical
6L     # integer, notice the 'L' at the end
-3.14  # double
"hi"   # character
```

```{r uncommon-atomic-types}
# complex
7 + 3i

# raw
charToRaw("abc123")
intToBits(2345)
```

## Non-Atomic Types

These are types that may contain more than one type of data. The two most common non-atomic types are `list` and `data.frame`.

```{r non-atomic-types}
list()
data.frame()
```

## Working with Types

### Checking the Type of an Object

We can check the type of an object by using the `typeof()` method.

```{r typeof-object}
typeof(TRUE)
typeof(3L)
typeof(6.1)
typeof("hello, world")
```

We can test for a specific type by using the `is.<type>()` family of functions.

```{r is-type}
is.logical(5)
is.double(pi)
is.integer(4.5)
is.character("I'm a string")
```

There are many more checks to use depending on the situation.

```{r is-type-special}
is.numeric(16)
is.atomic(5+4i)
is.function(sum)
is.nan(Inf - Inf)
```

Sometimes a number looks like an integer, but is actually a double:

```{r sneaky-double}
is.integer(4)
```

### Coercion and Conversion

We can convert the data type of one object to the type of another using the `as.<type>()` methods.

```{r conversion}
as.logical(7.8)
as.integer(4.0)
as.double(5L)
as.character(pi)
as.complex(3)
as.double("2.718")
```

What happens when we combine different types of data?

```{r mixing-data-types}
TRUE + 3L
5.5 + (4+5i)
```

What's going on here is something called *coercion*. This is when the type of one object is *implicitly* converted to another.

There are rules for coercion. Types are promoted to the *least restrictive type*. In general, the hierarchy is:

`logical > integer > double > character`

Can we add numbers and characters?

```{r numbers-and-characters}
5 + "4"
"1" + 1
```

Numbers and characters cannot be mixed in arithmetic, but numbers can still be coerced to characters. As we'll see soon, every atomic type is actually an *atomic vector.*

```{r number-coerced-to-character}
c(4, 5, "6")
```

Not every value can be coerced to every other type, or no robust checks will be performed to ensure validity of a conversion:

```{r inexact-conversion}
as.integer(4.5)
```

## Exercise

1.  Guess the output type:

```{r guess-output-type}
# a "regular" number
14
# integer division
14 %/% 3
# remainder
14 %% 3
# ceiling/floor/round
ceiling(pi)
round(pi)
```

2.  How can we more strongly check if a given double is an integer?

```{r check-double-is-integer}

```
