---
title: "Vectors, Lists, and Data Frames"
format: html
editor: visual
execute:
  eval: false
---

# Vectors

A *vector* is a one-dimensional collection of values of the same type. Vectors are the foundation of data in `R`. Everything in `R` is built upon vectors, making them a crucial concept to understand.

**Key Properties**

-   All elements must be of the same type
    -   elements will be *coerced* to the *least restrictive type*
-   Elements of the vector can be accessed by *indexing* using `[ ]`
-   They allow *vectorized operations*, meaning functions apply to all elements automatically

## Creating Vectors

Vectors can be created by:

**a)** using the *combine* function, `c()`

```{r create-with-combine}
c(4.5, 2, 3.8)
c(TRUE, FALSE, FALSE)
c("Apple", "Banana", "Cherry")

# what happens when you combine multiple vectors?
```

**b)** ranges and sequences

```{r create-range}
1:4
seq(from = 0.1, to = 1.0, by = 0.4)
```

**c)** built-in vectors

```{r built-in-vecs}
LETTERS
letters
```

## Indexing

You can extract specific elements using single *square brackets* `[ ]`.

```{r vector-indexing}
numbers <- c(10, 20, 30, 40, 50)

# get first element
numbers[1]

# get multiple elements
numbers[c(1, 3, 5)]

# get all except second element
numbers[-2]

# get all except specific elements
numbers[-c(1, 3)]

# access using logical conditions
numbers[c(TRUE, FALSE, FALSE, TRUE, TRUE)]
numbers[numbers > 20]
```

### Modifying a Value

Indexing can be combined with assignment.

```{r vector-index-assignment}
x <- 1:5
x

x[2] <- 7
x
```

## Vectorized Operations

R allows you to apply operations to an entire vector at once.

```{r vectorized-operations}
# arithmetic operations
numbers + 5
numbers * 2

# string operations
fruits <- c("Apple", "Banana", "Cherry")
paste(fruits, "Juice")

# logical operations
numbers > 25

# other functions
sin(numbers)
-numbers
10^(1:4)
(1:4)^10 # what happens without the parentheses?
```

## Functions for Vectors

`R` has built-in functions that operate on vectors.

```{r vector-functions}
length(numbers)
sum(numbers)
mean(numbers)
sort(c(5, 1, 3, 2, 4))
rev(fruits)
unique(c(0, 1, 1, 2, 3, 5))
```

## Coercion

A vector may only contain values of the same type.

```{r coercing-atomic-vector}
c(TRUE, 5, FALSE)
c(5, "6", 3 + 4i)
```

## Vector Exercises

Pick some random numbers

```{r pick-numbers}
# everyone pick a number between 1 and 10 inclusive
numbers <- c()

# count the number of occurrences of each pick
table(numbers)
```

Explore a uniform distribution

```{r uniform-numbers}
# construct a U(0, 10) distribution

# how many numbers are there?

# how many numbers are less than 6?

# what proportion of the numbers are less than 6?

# what is the sum of all the numbers?

# what is the mean of all the numbers?
```

# Matrices

A *matrix* is a 2-dimensional rectangular structure where all the elements are of the same type (just like a vector, but with rows and columns).

**Key Properties**

-   Matrices have rows and columns

-   All elements must be of the same type

    -   elements will be *coerced* to the *least restrictive type*

-   Can be created using the `matrix()` function

-   Indexing uses `[row, column]`

## Creating a Matrix

A matrix can be created with `matrix()`

```{r creating-a-matrix}
my_matrix <- matrix(1:9, nrow = 3, ncol = 3)
```

Notice that the matrix is filled *column-wise* by default

```{r show-column-wise-fill}
my_matrix
```

To fill the matrix *row-wise*, use `byrow=TRUE`

```{r row-wise-matrix}
my_matrix <- matrix(1:9, nrow = 3, ncol = 3, byrow = TRUE)
my_matrix
```

**Questions**

1.  Do we need to specify both the number of rows and number of columns?
2.  What if the data is smaller than the matrix?
3.  What if the data is larger than the matrix?

## Indexing

Matrices use row and column indexing, `[row, column]`

```{r matrix-indexing}
# get element in row 2, column 3
my_matrix[2, 3]

# get entire second row
my_matrix[2, ]

# get entire first column
my_matrix[, 1]
```

We can also index multiple rows and columns.

```{r matrix-indexing-multiple}
# rows 1 and 3 - columns 2 and 3
my_matrix[c(1, 3), c(2, 3)]

# first row, twice - columns 1 and 3
my_matrix[c(1, 1), c(1, 3)]
```

If the matrix is a single dimension, then `R` will automatically *drop the dimension.*

```{r matrix-drop-dim}
# dimension dropped to vector
my_matrix[1, ]
# force R to keep dimension
my_matrix[1, , drop = FALSE]
```

## Matrix Operations

In `R`, Matrices are just vectors in disguise, thus they support vectorized operations in the same way that vectors do. I.e., operations are applied *element-wise*.

```{r matrix-operations}
my_matrix * 2

matrix_A <- matrix(1:4, nrow = 2)
matrix_B <- matrix(5:8, nrow = 2)
matrix_A + matrix_B

# element-wise multiplication
matrix_A * matrix_B

# true matrix multiplication


# square each element
my_matrix^2
```

## Functions for Matrices

There are some functions that are designed to specifically work on matrices.

```{r matrix-functions}
# getting the dimensions
nrow(my_matrix)
ncol(my_matrix)

# row and column sums
rowSums(my_matrix)
colSums(my_matrix)

# row and column means
rowMeans(my_matrix)
colMeans(my_matrix)

# transpose
t(my_matrix)
```

Other functions can be applied to the rows or columns of a matrix.

```{r apply-to-matrix}
# apply function to rows
apply(my_matrix, MARGIN = 1, FUN = sum)

# apply function to columns
apply(my_matrix, MARGIN = 2, FUN = mean)
```

## Matrix Exercises

```{r matrix-exercise}
# create a matrix of letters
letter_matrix <- matrix()

# get the diagonal elements of this matrix
m <- matrix(1:16, nrow = 4)
```

# Arrays

An *array* is an extension of a matrix, and supports more than two dimensions.

**Key Properties**

-   Arrays have multiple dimensions

-   All elements must be of the same type

    -   elements will be *coerced* to the *least restrictive type*

-   Can be created using the `array()` function

-   Indexing uses `[i, j, k, ...]`

-   Supports vectorized operations

## Creating an Array

An array can be created with `array()`

```{r creating-an-array}
my_array <- array(1:18, dim = c(3, 3, 2))
my_array
```

A 3-dimensional array can be thought of as layers of matrices, where the order is `[row, column, layer]`.

```{r layers-of-matrices}
# get the second matrix from the array
my_array[, , 2]
```

## Reshaping an Array

Arrays (and matrices and vectors) can be reshaped into any other dimension by using the `dim()` method with assignment.

```{r reshape-array}
dim(my_array) <- c(2, 3, 3)
my_array
```

## Array Operations and Functions

Operations and functions work on arrays element-wise, just like matrices and vectors.

```{r array-operations}
3 * my_array

array_A <- array(2 * (1:18), dim = c(3, 3, 2))
array_B <- array(3 * (1:18), dim = c(3, 3, 2))
array_B - array_A
```

## Array Exercises

What happens if we apply operators to arrays of different dimensions?

```{r dimension-mismatch}
```

What order does an array get filled with values?

```{r array-fill-order}
```

Get the row means of each layer of a $2 \times 3 \times 5$ array. Predict what the shape/size/length of the result will be.

```{r array-row-means}
```
