---
title: "Lists and Data Frames"
format: html
editor: visual
execute:
  eval: false
---

# Lists

A *list* in `R` is a versatile data structure that can store multiple data types in a single object. Unlike vectors, matrices, and arrays (which require all elements to be of the same type), lists can hold heterogeneous data—numbers, characters, vectors, and even other lists.

**Key Properties**

-   Can store objects of different types

-   Elements can be named for better organization

-   Supports nested structures (lists of lists of lists)

**Related objects in other languages**

-   Python - `Dictionary`

-   Julia - `Dict{Any, Any}`

-   C# - `Dictionary<TKey, TValue>`

-   C++ - `std::map<TKey, TValue>`

## Creating a List

Lists can be created using the `list()` function.

```{r creating-a-list}
list(1, "b", 5.4)
list(1:5, letters[1:4], c(3.14, 2.718))
list("a", list(1, 2), 7)
```

Elements of the list can be named for easier access.

```{r naming-elements}
list(a = 1:4, b = c("a", "b", "c"))
list(letters = letters[1:5], x = 1:10)

my_list <- list(
  name = "Alice",
  age = 25,
  scores = c(90, 85, 88),
  passed = TRUE
)
my_list
```

## Indexing

Indexing a single element of a list can be done in two ways

**1)** With double brackets `[[ ]]`

```{r double-bracket-index}
# element at index 3
my_list[[3]]

# element with name "age"
my_list[["age"]]
```

**2)** With the `$` for named objects

```{r named-object-index}
my_list$name
```

Accessing using the `$` is convenient because you can use tab completion to get a list of named elements.

### Sub-Lists

Lists can be indexed using single brackets `[ ]` to create *sub-lists*.

```{r creating-sublists}
my_list[1]
my_list[2:3]
my_list[c("name", "scores")]
```

### Indexing Summary

| Method | Example | Description |
|----|----|----|
| `[[ ]]` | `my_list[[2]]`, `my_list[["age"]]` | Gets the *element* at the index |
| `$` | `my_list$name` | Gets the *element* with the name |
| `[ ]` | `my_list[3:4]`, `my_list[c("age", "name")]` | Gets a *list* containing the elements |

## Modifying a List

Elements of a list can be added or modified.

```{r modifying-a-list}
my_list <- list(age = 26)

# changing a value
my_list$age <- 30

# adding new values
my_list$numbers <- 1:3
my_list$letters <- letters[5:7]


my_list
```

What happens if we add a value by index?

```{r adding-value-by-index}
my_list <- list()
my_list[[3]] <- 1
my_list
```

We can remove values by setting them to `NULL`

```{r removing-elements}
my_list <- list(
  name = "Alice",
  age = 25,
  scores = c(90, 85, 88),
  passed = TRUE
)

my_list$age <- NULL

my_list
```

Elements can be appended to a list

```{r append-to-list}
new_list <- append(my_list, list(city = "New York"))
```

And lists can be combined

```{r combine-lists}
list1 <- list(a = 1, b = 2)
list2 <- list(c = 3, d = 4)
c(list1, list2)
```

## Nested Lists

Lists can contain other lists.

```{r nested-lists}
nested_list <- list(
  person = list(name = "Bob", age = 30),
  scores = list(math = 85, science = 90)
)

nested_list$person$name
nested_list$scores$science
```

## Unlisting a List

Lists can be *unlisted* to a regular vector.

```{r unlisting}
my_list <- list(1, "2", 3 + 4i)
unlist(my_list)
```

**Caution:** If a list has mixed types (numeric, character, logical), then `unlist()` converts everything to the least restrictive type.

## Functions for Lists

There are a few methods that work directly on lists.

```{r functions-on-lists}
my_list <- list(
  name = "Alice",
  age = 25,
  scores = c(90, 85, 88),
  passed = TRUE
)

length(my_list)
names(my_list)
unlist(my_list)
```

Functions can be applied to each element in a list with the `lapply()` function, which stands for *list apply*.

```{r lapply}
lapply(my_list, FUN = length)
```

There are other `*apply()` methods as well:

-   `sapply()` - works like `lapply()` but it trues to simply the output into a vector

-   `vapply()` - similar to `sapply()`, but more strict about the expected output type

## Filtering Lists

You can use the `Filter()` function to return only the elements that match a condition

```{r filter-a-list}
num_list <- list(a = 1:5, b = 6:10, c = 11:15)

# keep only elements where the mean is greater than 5
Filter(function(x) mean(x) > 5, num_list)
```

## Recap

-   Lists are generic vectors

-   Elements can be any kind of object, even other lists

-   Accessing elements of a list can be done with `[[ ]]` or `$`

-   Creating a sublist can be done with `[ ]`

-   Elements of a list can be named

-   Functions can be applied to the elements of a list with the `*apply()` family of functions

-   Elements of a lists can be filtered by a condition

# Data Frames

**Describe a Data Frame**

-   

-   

-   

## Creating a Data Frame

A data frame can be created using the `data.frame()` method, similar to how you construct a list.

```{r create-a-dataframe}
df <- data.frame(
  Name = c("Alice", "Bob", "Charlie"),
  Age = c(25, 30, 22),
  Score = c(90, 85, 88)
)

df
```

We can verify some properties of a data frame

```{r dataframe-properties}
is.data.frame(df)
is.list(df)
typeof(df)
class(df)
unclass(df)
str(df)
```

## Indexing

Because a data frame is just a list, we can access elements (columns) using both `[[ ]]` and `$`.

```{r dataframe-indexing}
df[["Name"]]
df$Age
df[[3]]
```

To get a new data frame with select columns, you can use single brackets `[ ]`

```{r dataframe-indexing-single-bracket}
df[3]
df["Score"]
df[1:2]
df[sort(names(df))]
df[c(1, 1, 1, 1)]
df[-2]
```

You can also index into a data frame in the same way as a matrix. E.g., `[rows, cols]`

```{r df-row-col-index}
df[, "Age"]
df[, "Age", drop = FALSE]
df[c(1, 3), c(1, 3)]
df[2:3, ]
df[c(TRUE, FALSE, FALSE), ]
```

**Questions**

1.  Under what conditions is a vector returned?
2.  Under what conditions is a data frame returned?

## Subsetting

While we can use `[ ]` to get select rows and columns, we can also use a more readable `subset()` function to filter the rows.

```{r dataframe-subset}
subset(df, Score > 85)
```

What is the equivalent `[ ]` expression?

```{r dataframe-subset-using-bracket}
```

## Modifying a Data Frame

We can add new columns to a data frame

```{r adding-columns}
df$Passed <- df$Score > 80
df
```

## Functions for Data Frames

Since a data frame is just a list, we can use `lapply()` and `sapply()`

```{r df-lapply}
lapply(df[, 2:3], mean)
sapply(df[, 2:3], mean)
```

## Exercises

Can we automatically select just the numeric columns and apply the function?

```{r filter-and-transform}
```

How does *vector recycling* work for data frames?

```{r vector-recycling}
```

How can we filter with multiple conditions?

```{r filter-multiple-conditions}
```

Practice selecting, filtering, and modifying using the `mtcars` data set.

```{r practice-with-mtcars}
df <- mtcars
str(df)
```
