---
title: "Reading and Visualizing Data"
format: html
editor: visual
execute:
  eval: false
---

# Reading Data

We can read in data in several ways - some automated and some manually. Before reading the data, it's a good idea to view its structure.

## Viewing the Data

RStudio supports viewing a data set before importing it. When you click on a data file in the `Files` tab of the lower right pane, you will get an option to view or import the data.

## Importing via RStudio

There are several ways to import data using RStudio.

1.  `File > Import Dataset`
2.  `Environment > Import Dataset`
3.  `Files > Click on file`

## Importing Manually

Structured text files like CSV files can be read using the `read.csv()` method.

```{r read-csv}
df <- read.csv("../data/simple_csv.csv")
```

Other read functions include `read.table()`, `read.csv2()` , `read.delim()` , and `read.delim2()`. All of these are different variants of `read.table`, just with different defaults. Use `?read.table` to get an idea of all the possible controls for reading delimited files.

## Reading Excel Files

Reading Excel files can be achieved by using the `readxl` library, and it can be installed and automatically used by using the built in `Import Dataset` functionality of RStudio.

```{r read-xlsx}
```

## Reading Other Data Types

Many fields will use other data formats which cannot be read using a built in `R` method. In those cases, a package may exist to help read in the data.

# Visualizing Data

R has built-in support for creating graphs of different types.

## Simple Plots

There are a few simple plotting functions in R

-   `plot()`
    -   also "generic"
    -   can produce scatter and line plots
-   `lines()` adds lines to plots
-   `points()` adds points to plots
-   `curve()` plots a function over a domain
    -   can be used alone or "layered" on top of an existing plot

Other types of graphs include

-   `hist()` for histograms
-   `barplot()`
-   `boxplot()`
-   `pie()` for pie charts
-   `stem()` for stem-and-leaf plot

**Scatter**

```{r scatter-plot}
plot(mtcars$wt, mtcars$mpg,
  main = "Scatter plot of MPG vs Weight",
  xlab = "Weight",
  ylab = "Miles per Gallon",
  pch = 19,
  col = "blue"
)
```

**Lines**

```{r line-plot}
plot(AirPassengers,
  main = "Monthly Air Passengers",
  xlab = "Year",
  ylab = "Number of Passengers",
  col = "darkgreen",
  type = "o",
  pch = 16
)
```

**Bar Plot**

```{r bar-plot}
barplot(table(mtcars$cyl),
  main = "Car Count by Number of Cylinders",
  xlab = "Number of Cylinders",
  ylab = "Count",
  col = "lightblue",
  border = "black"
)
```

**Histogram**

```{r histogram-plot}
hist(iris$Sepal.Length,
  main = "Histogram of Sepal Length",
  xlab = "Sepal Length",
  col = "skyblue",
  border = "white"
)
```

**Density**

```{r density-plot}
plot(density(iris$Sepal.Length),
  main = "Density Plot of Sepal Length",
  xlab = "Sepal Length",
  col = "blue"
)
```

**Box Plot**

```{r box-plot}
boxplot(Sepal.Length ~ Species,
  data = iris,
  main = "Boxplot of Sepal Length by Species",
  xlab = "Species",
  ylab = "Sepal Length",
  col = c("lightblue", "lightgreen", "lightcoral")
)
```

**Pie Chart**

```{r pie-chart}
pie(table(mtcars$cyl),
  main = "Distribution of Cars by Number of Cylinders",
  col = c("lightblue", "lightgreen", "lightcoral")
)
```

## Plotting Curves

Curves can be plotted using the `curve()` method.

```{r plotting-a-curve}
curve(sin(x),
  from = -10, to = 10, col = "blue",
  main = "Plot of sin(x)",
  xlab = "x", ylab = "sin(x)"
)
```
