---
title: "Data Wrangling with dplyr and tidyr"
format: html
editor: visual
execute:
  eval: false
---

```{r setup}
library(tibble)
library(dplyr)
library(tidyr)
```

# The `tidyverse`

The `tidyverse` is a collection of R packages designed for data science that all share an underlying philosophy and common design principles.

**Key Packages**

-   `tibble` for modern data frames with enhanced functionality

-   `readr` for reading and writing data files

-   `dplyr` for data manipulation and transformation (e.g., filtering, summarizing)

-   `tidyr` for tidying and reshaping data

-   `ggplot2` for data visualization

-   `purrr` for functional programming (e.g., working with lists and applying functions)

-   `stringr` for string manipulation

-   `forcats` for working with categorical variables (factors)

## Tibbles

A `tibble` is a modern version of a data frame in `R`, and is the default data frame type in the `tidyverse`. Tibbles are compatible with data frames, but have a few enhancements.

### Compared to a Data Frame

Tibbles display only the first 10 rows and the columns that fit on the screen. Data frames, print the entire dataset by default.

```{r printing-a-tibble}
as.data.frame(mtcars)
as_tibble(mtcars) # where did the row names go?
```

Tibbles do not automatically convert character vectors into factors, whereas `R` prior to version 4 used to read strings as factors by default.

```{r strings-as-factors}
# old behavior
read.csv("../data/billboard.csv", stringsAsFactors = TRUE)
# new behavior
read.csv("../data/billboard.csv")
# tibble returned by readr
readr::read_csv("../data/billboard.csv")
```

When selecting a single column from a tibble, another tibble is returned. A data frame may return a vector when a single column is selected.

```{r tibble-single-column}
df <- mtcars
tbl <- as_tibble(mtcars, rownames = "make-model")

df[, "mpg"]
tbl[, "mpg"]
```

Tibbles allow non-syntactic column names (e.g., names with spaces or starting with numbers), while data frames require column names to follow R's naming conventions.

```{r tibble-column-names}
data.frame("Column 1" = 1:3, "123column" = 4:6)
tibble("Column 1" = 1:3, "123column" = 4:6)
```

Tibbles are strict; they complain more often which can help catch bugs. E.g., if you try to create a tibble with mismatched column lengths, it will throw an error. Prior to version 4.0, a data frame would automatically recycle values.

```{r tibble-strictness}
data.frame(A = 1:3, B = 1:2)
tibble(A = 1:3, B = 1:2)
```

## dplyr

`dplyr` provides a set of functions to manipulate data frames and tibbles. From the docs:

> `dplyr` is a grammar of data manipulation, providing a consistent set of verbs that help you solve the most common data manipulation challenges:
>
> -   `mutate()` adds new variables that are functions of existing variables
>
> -   `select()` picks variables based on their names
>
> -   `filter()` picks cases based on their values
>
> -   `summarise()` reduces multiple values down to a single summary
>
> -   `arrange()` changes the ordering of the rows

## tidyr

**`tidyr`** helps reshape data, making it "tidy"—that is, ensuring each variable is a column, each observation is a row, and each value is a cell.

The [tidyr docs](https://tidyr.tidyverse.org/index.html) are a great resource for learning about tidy data, especially the vignette on [tidy data](https://tidyr.tidyverse.org/articles/tidy-data.html). From the vignette:

> Tidy datasets and tidy tools work hand in hand to make data analysis easier, allowing you to focus on the interesting domain problem, not on the uninteresting logistics of data.

## Non-Standard Evaluation (NSE)

One last point before we begin, tidyverse verbs often use *non-standard evaluation*. Without going into too much detail, this allows you to use the names of columns **unquoted** inside of verbs.

```{r non-standard-evaluation}
mtcars %>% select(mpg, hp)
```

# Data Wrangling

Introducing the pipe operator `%>%` provided by the `magrittr` package and made available by almost every other `tidyverse` package. It takes the left hand side and inserts it into the function on the right.

```{r pipe-operator}
# simple pipe
3.14 %>% sin()
0.717 %>% cos()

# can put RHS into any location using '.' placeholder
5 %>% seq(from = 1, to = .)

# can chain pipes for easier reading
mtcars %>%
  group_by(cyl) %>%
  select(cyl, mpg, disp, hp) %>%
  summarize(across(where(is.double), mean))
```

As of `R 4.1` there is a native pipe operator, `|>`. It largely works the same way as `%>%`. One can argue that `%>%` is better because it can be used in versions of R `<4.1`, thus being more compatible with older systems.

I will not speak to which is better, but I will continue to use `%>%` because of muscle memory.

## dplyr Verbs - Select, Filter, Mutate

### **Selecting Columns**

Use `select()` to keep or remove specific columns.

```{r selecting-columns}
# Select only mpg, cyl, and hp columns

# Remove the wt and qsec columns

# Select columns using a range
```

**Exercise:** Select only the `mpg`, `hp`, and `gear` columns from `mtcars`.

```{r select-exercise}
```

### **Filtering Rows**

Use `filter()` to subset rows based on one or more conditions.

```{r filtering-rows}
# Select cars with mpg greater than 25

# Select cars with 4 cylinders and mpg greater than 20

# Select cars with either 4 or 6 cylinders
```

**Exercise:** Find all cars with horsepower greater than 150 and weight less than 3.

### **Creating New Columns**

Use `mutate()` to add new variables or modify existing ones.

```{r mutating-columns}
# Create a new column for power-to-weight ratio

# Convert weight from tons to kilograms
# 453.592 kg per ton
```

**Exercise 1:** Create a new column that categorizes cars as `"High MPG"` if `mpg > 20` and `"Low MPG"` otherwise using `mutate()` and `ifelse()`.

**Exercise 2**: Write a function that classifies a car as `"High MPG"`, `"Med MPG"`, and `"Low MPG"` if the MPG is `>=25`, `>=20`, and `<20` respectively.

### **Sorting Rows by Columns**

Use `arrange()` to order rows by one or more variables.

```{r sorting-rows}
# Arrange data in ascending order of mpg

# Arrange data in descending order of horsepower

# Arrange by multiple columns: mpg (ascending), then hp (descending)
```

**Exercise:** Sort `mtcars` by `cyl` (ascending) and `mpg` (descending). 

## Grouping and Summarizing

Use `group_by()` to group data by one or more variables, and `summarize()` to compute summary statistics. The "group by and summarize" pattern is very common.

```{r group-by-and-summarize}
# Calculate average mpg for each cylinder group

# Count the number of cars in each cylinder group

# Calculate the average mpg and max horsepower for each cylinder
```

Note: I like to have one verb per line to make my code more readable.

## Other dplyr Verbs

### **Counting Rows**

Use `count()` as a shortcut for `group_by()` + `summarize(n())`.

```{r counting-rows}
# Count occurrences of each cylinder category

# Count occurrences of each gear category
```

**Exercise:** Count the number of cars for each `cyl` and `gear` combination.

```{r counting-exercise}
```

### **Removing Duplicates**

Use `distinct()` to get unique rows based on selected columns.

```{r distinct-rows}
# Get unique cylinder values

# Get unique combinations of cyl and gear
```

**Exercise:** Find the unique combinations of `cyl` and `am` (automatic/manual transmission).

```{r distinct-exercise}
```

### **Renaming Columns**

Use `rename()` to give columns more meaningful names. This can be especially helpful when polishing a table for publication.

```{r renaming-columns}
# Rename mpg to Miles_per_Gallon

# Rename multiple columns
```

**Exercise:** Rename `disp` to `"Displacement"` and `drat` to `"Rear_Axle_Ratio"`.

```{r renaming-exercise}
```

### **Relocating Columns**

Use `relocate()` to move columns to a new position.

```{r relocating-columns}
# Move mpg to the first column

# Move hp before mpg
```

**Exercise:** Move the `gear` column to be the first column.

```{r relocating-exercise}
```

## Putting it all Together

```{r all-together}
mtcars %>%
  filter(mpg > 20) %>%
  select(mpg, cyl, hp, wt) %>%
  mutate(power_to_weight = hp / wt) %>%
  arrange(desc(power_to_weight))
```

## Reshaping

`tidyr` is used for reshaping and cleaning datasets. It helps transform data between *wide* and *long* formats, handle missing values, and make datasets "tidy"—where each variable has its own column, each observation has its own row, and each value has its own cell.

**Key Functions**

-   `pivot_longer()`

-   `pivot_wider()`

-   `separate()`

-   `unite()`

### **Pivot Longer**

Used when multiple columns contain the same type of information. We need to gather them into two columns (one for variable names, one for values).

```{r pivot-longer}
# sales in each year among different groups
tbl <- tibble(
  id = c(1, 2, 3),
  year_2020 = c(100, 200, 300),
  year_2021 = c(150, 250, 350)
)

# "year" information in the columns
# year needs to be its own column
# value needs to be another column
```

### **Pivot Wider**

Used to spread values across multiple columns. The opposite of `pivot_longer()`.

```{r pivot-wider}
finances <- tibble(
  id = c(1, 1, 2, 2),
  year = c("2020", "2021", "2020", "2021"),
  sales = c(100, 150, 200, 250)
)

# put into "wider" format by year
```

### **Separating Columns**

Sometimes the data in a column can be split up into new columns. This can be done using the `separate()` function.

```{r separate-column}
# Example dataset
data <- tibble(
  full_date = c("2023-01-01", "2023-02-01", "2023-03-01")
)

# Separate the column into year, month, and day
```

### **Unite Columns Together**

Opposite of separating columns is `unite()` which joins the data of two columns together into a single column.

```{r unite-columns}
data <- tibble(
  year = c("2023", "2024"),
  month = c("01", "02"),
  day = c("15", "20")
)

# Combine into a full_date column
```

## Handling NA Values

`tidyr` also supports handling missing (`NA`) values.

-   `drop_na()`

-   `replace_na()`

-   `fill()`

```{r handling-missing-data}
tb <- readr::read_csv("../data/tb.csv")
glimpse(tb)

# see how much of the data is missing
tb %>%
  summarize(across(everything(), ~ mean(is.na(.)))) %>%
  pivot_longer(
    everything(),
    names_to = "Column",
    values_to = "Prop. NA"
  )

# select the top 5 most represented columns and drop NA
```

## Joining Datasets

Data can come from multiple sources, and it is a common task to join the data into a single dataset. There are several types of joins that are intended for different scenarios:

-   `left_join()`

-   `right_join()`

-   `inner_join()`

-   `full_join()`

-   `semi_join()`

-   `anti_join()`

Let's look at some examples using a made up data relating employees to departments.

```{r join-data}
employees <- tibble(
  emp_id = c(1, 2, 3, 4, 5),
  name = c("Alice", "Bob", "Charlie", "David", "Ellie"),
  dept_id = c(101, 102, 103, 104, 101)
)

departments <- tibble(
  dept_id = c(101, 102, 105),
  dept_name = c("HR", "Finance", "IT")
)

employees
departments
```

Note: See the `dplyr` [cheat sheet](https://github.com/rstudio/cheatsheets/blob/main/data-transformation.pdf) for a handy infographic detailing the various joins.

### **Left Join**

Keeps all rows from the left dataset (`employees`), adding matching columns from `departments`.

```{r left-join}
```

### **Right Join**

Keeps all rows from the right dataset (`departments`), adding matching columns from `employees`.

```{r right-join}
```

### **Inner Join**

Keeps only rows where there is a match in both datasets.

```{r inner-join}
```

### **Full Join**

Keeps all rows from both datasets, filling missing values with `NA`.

```{r full-join}
```

### Which Join Should You Use?

-   Use **`left_join()`** when your primary dataset is on the left and you don’t want to lose any rows

-   Use **`inner_join()`** when you only care about rows that have a match in both datasets

-   Use **`full_join()`** when you want to retain all information from both datasets

-   Use **`semi_join()`** if you just want to filter `employees` to those that match `departments`

-   Use **`anti_join()`** to find rows in `employees` that *don’t* have a match in `departments`
