---
title: "Data Visualization with ggplot2"
format: html
editor: visual
execute:
  eval: false
---

```{r setup}
library(tidyverse)
```

# Introduction to `ggplot2`

`ggplot2` is built on the *grammar of graphics*, which is a structured framework to create statistical graphs.

**Key Concepts**

The grammar of graphics breaks down a graph into fundamental components:

-   Data – The dataset being visualized

-   Aesthetics – The mapping between data variables and visual properties

-   Geometries – The type of geometric object used to represent data

-   Scales – Transformations that map data values to aesthetic attributes

-   Statistics – Statistical transformations applied to the data before plotting (e.g., smoothing, binning)

-   Coordinates – The coordinate system used

-   Facets – The method of splitting data into subplots (small multiples) for comparison

The library `ggplot2` follows a simple formula:

```         
ggplot(data = <data>, aes(x = <col1>, y = <col2>)) +
  geom_*()
```

The aesthetics (`aes`) can be specified on whole-plot basis (within the `ggplot()` function) or on a layer basis (`geom_*(aes(x = <col1>, y = <col2>))`).

## Common Geoms

### Scatter Plots

```{r geom-point}
# plot the "mtcars" data of mpg vs hp
```

We can control the size, shape, and color of the points. We can even set the color according to other variables.

```{r geom-point-colored}
ggplot(mtcars, aes(x = hp, y = mpg, color = factor(cyl), size = wt)) +
  geom_point()
```

### Bar Charts

```{r bar-chart}
# plot the "diamonds" data of cut quality
```

We can change the `y`-axis to show proportions by computing the proportion and using `geom_col()`.

```{r bar-chart-proportion}
diamonds %>%
  group_by(cut) %>%
  summarise(proportion = n() / nrow(diamonds)) %>%
  ggplot(aes(x = cut, y = proportion)) +
  geom_col()
```

### Line Plots

```{r line-plots}
# plot the "economics" data of unemployment over time
```

### Box Plots

```{r box-plots}
# plot the "mtcars" data of mpg vs cylinders
```

### Histogram and Density Plots

## Customizing the Plots

### Adding Titles, Labels, and Themes

```{r titles-labels-themes}
# add titles, axis labels, and change the theme
ggplot(mtcars, aes(x = hp, y = mpg, color = factor(cyl))) +
  geom_point()
```

### Faceting

```{r faceting}
ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point() +
  facet_wrap(~cyl)
```

### Colors and Scales

```{r colors-and-scales}
ggplot(mtcars, aes(x = hp, y = mpg, color = factor(cyl))) +
  geom_point() +
  scale_color_brewer(palette = "Set1")
```

# Exploring Datasets

Let's take some time to explore some of the built-in datasets by creating exploratory and explanatory plots.

**Recommended Datasets**

| Dataset | Description | Best for Visualizations |
|-----------------|----------------------------|----------------------------|
| `gapminder` | Country-level economic data | GDP, Life Expectancy, Population trends |
| `mpg` | Car fuel efficiency | Engine size, MPG comparisons |
| `diamonds` | Diamond prices & attributes | Price trends, Cut quality comparisons |
| `iris` | Flower measurements | Sepal vs Petal dimensions, Species classification |
| `economics` | US Economic data | Unemployment trends, Population growth |
| `airquality` | New York air pollution (1973) | Temperature vs Ozone, Time series trends |
| `ChickWeight` | Chicken growth under diets | Weight gain comparisons over time |
| `penguins` | Penguin species & body sizes | Species-based comparisons, Bill length vs depth |
