---
title: "Applying tidyverse and ggplot2 in Real-World Scenarios"
format: html
editor: visual
execute:
  eval: false
---

```{r setup}
library(tidyverse)
```

# Tidying Messy Data

Five most common problems of messy datasets:

1)  Column headers are values, not variable names
2)  Multiple variables are stored in one column
3)  Variable names are stored in both rows and columns
4)  Multiple types of observational units are stored in the same variable
5)  A single observational unit is stored in multiple tables

A small set of tools can be used to clean most of the problems of messy data:

1)  Melting (pivot longer)
2)  String manipulation (splitting)
3)  Casting (pivot wider)

## Guided Example - Billboard Top Hits

### Reading the Data

Let's start by reading in the data using `readr::read_csv().` I'm also going to apply some quick fixes before we begin tidying. Specifically renaming a column, pre-processing `artist`, `time`, and `track`, and arranging the rows by `year`.

```{r reading-billboard}
df0 <- readr::read_csv("../data/billboard.csv",
  col_types = cols(time = col_character())
) %>%
  select(
    year, artist.inverted, track, time, genre,
    date.entered, x1st.week:x76th.week
  ) %>%
  rename(artist = artist.inverted) %>%
  mutate(
    artist = iconv(artist, "MAC", "ASCII//translit"),
    time = lubridate::ms(time),
    track = stringr::str_replace(track, " \\(.*?\\)", "")
  ) %>%
  arrange(year, artist, track)
```

-   What can we do to examine the structure of the data?
-   What are the column names?
-   What are the column types?
-   How many rows and columns are there?
-   Is there any missing data?

```{r initial-questions}
```

Is the data in a tidy format? What is wrong?

-   \<answers\>

### Fixing the Column Names

The column names are a bit messy. We can rename them into something a little nicer looking and easier to look at.

```{r fixing-week-columns}
# manual fix
names(df0)[-c(1:5)] <- str_c("wk", 1:76)
```

Below is a more "tidy" way of fixing the names.

```{r fixing-week-columns-regex}
fix_week_name <- function(name) {
  return(paste0("wk", str_extract(name, "\\d+")))
}

df0 <- df0 %>%
  rename_with(fix_week_name, .cols = ends_with("week"))
```

### Putting the Data into a Tidy Format

The ranks for each song are stored in multiple `week` columns. We'll need to pivot the data into a longer format.

```{r pivot-billboard-longer}
```

Pre-coded chunk for reference:

```{r pivot-billboard-longer-precoded}
df1 <- df0 %>%
  pivot_longer(
    cols = wk1:wk76,
    names_to = "week",
    values_to = "rank",
    values_drop_na = TRUE
  ) %>%
  mutate(
    week = str_remove(week, "wk"),
    week = as.integer(week)
  )
```

### Splitting into Songs and Ranks

We can split the data into two datasets: songs and ranks. This is so that each dataset represents a single thing, and they can be joined back together later if needed.

```{r split-songs}
# select artist, track, and time
# get unique rows
# add a unique id (row number)
```

We can join the songs back with the dataset and then select just the `date.entered` and `rank` columns.

```{r split-ranks}
# join songs with dataset
# select the date.entered and rank
```

Pre-coded chunk for reference:

```{r split-clean-join-precoded}
song <- df1 %>%
  select(artist, track, time, year, genre) %>%
  distinct() %>%
  mutate(id = row_number()) %>%
  relocate(id, .before = 1)

rank <- left_join(df1, song) %>%
  relocate(id, .before = 1) %>%
  select(id, date.entered, rank, week)
```

We now have a tidy dataset! Here is the re-combined data.

```{r recombined-data}
df_full <- inner_join(song, rank, by = "id")
```

# Visualizing our Tidy Data

**How Many Weeks Do Songs Stay on the Chart?**

```{r}
ggplot(df_full, aes(x = week)) +
  geom_histogram(binwidth = 2, fill = "lightblue", color = "black") +
  labs(
    title = "Distribution of Weeks on the Billboard Chart",
    x = "Weeks on Chart",
    y = "Number of Songs"
  )
```

**Most Popular Artists Based on Chart Appearances**

```{r}
df_full %>%
  count(artist, sort = TRUE) %>%
  top_n(10) %>%
  ggplot(aes(x = reorder(artist, n), y = n, fill = artist)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 10 Artists with Most Billboard Chart Appearances",
    x = "Artist",
    y = "Number of Songs on Chart"
  )
```

**Rank Distribution by Genre**

```{r}
ggplot(df_full, aes(x = rank, fill = genre)) +
  geom_density(alpha = 0.5) +
  scale_x_reverse() +
  labs(
    title = "Rank Distribution by Genre",
    x = "Billboard Rank (1 = Top)",
    y = "Density"
  )
```

**Comparing Weeks on Chart by Genre**

```{r}
ggplot(df_full, aes(x = genre, y = week, fill = genre)) +
  geom_boxplot() +
  coord_flip() +
  labs(
    title = "Weeks on Chart by Genre",
    x = "Genre",
    y = "Weeks on Chart"
  )
```
